Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-22 mpaper/paper.tex
Cleaned up the second section

@@

The optimal \emph{policy} $\pi : \mathcal{S} \to \mathcal{A}$ (i.e., a choice of
actions for each state that maximises reward over time) is usually constructed
by defining a \emph{value (utility) function} $\V : \mathcal{S} \to \mathbb{R}$
that measures how good a state is based on the reward $\mathbf{r}$ as well as
the structure of the MDP. One can then find $\V$ by applying the Bellman backup
operator until convergence to every $s \in \mathcal{S}$ (the technique is known
as \emph{value iteration}) \cite{DBLP:books/daglib/0023820}:
\[
  V_{\mathbf{r}}(s) \coloneqq r(s) + \gamma \max_{a \in \mathcal{A}} \sum_{s' \in
    \mathcal{S}} \mathcal{T}(s, a, s')V_{\mathbf{r}}(s').
\]

However, we follow previous work on GP IRL
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}, and use a
\emph{linearly solvable} (or \emph{maximum causal entropy}) MDP with stochastic
policy that define probability distributions over actions (instead of
suggesting a single action for each state) \cite{ziebart2008maximum}. This type
of MDP can be solved by applying the 'soft' version of the operator
\cite{DBLP:conf/nips/LevinePK11,supplementary_material}:
\begin{equation} \label{eq:update_rule}
  \V(s) \coloneqq \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s'
      \in \mathcal{S}} \mathcal{T}(s, a, s')\V(s') \right).
\end{equation}
With this model, we can express the likelihood as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{equation} \label{pDr}
  \begin{split}
    p(\mathcal{D} \mid \mathbf{r}) &= \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} \mid s_{i,t}) \\
    &= \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right),
  \end{split}
\end{equation}
where
\[
  Q_{\mathbf{r}}(s, a) = r(s) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s, a, s')\V(s').
\]

However, a reward function learned by maximising this likelihood is not
transferable to new situations
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}. One needs to model the
reward structure in a way that would allow reward predictions for previously
unseen states.

One way to model rewards without assumptions of linearity is with a
\emph{Gaussian process} (GP). A GP is a collection of random variables, any
finite combination of which has a joint Gaussian distribution
\cite{DBLP:books/lib/RasmussenW06}. We write $r \sim \mathcal{GP}(0,
k)$ to say that $r$ is a GP with mean $0$ and covariance function
$k$. \emph{Covariance functions} (also known as \emph{kernels}) take two state
feature vectors as input and quantify how similar the two states are, in a sense
that we would expect them to have similar rewards.

As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states. Then the
full joint probability distribution can be factorised as
\begin{equation} \label{eq:full}
  \pfull = p(\mathbf{u}) \times p(\mathbf{r} \mid \mathbf{u}) \times
  p(\mathcal{D} \mid \mathbf{r}),
\end{equation}
where
\begin{align*}
  p(\mathbf{u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \\
                &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \\
                &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right)
\end{align*}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, and $m \in \mathbb{N}$ is
the number of inducing points. The GP posterior is then a multivariate Gaussian
\cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} \mid \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} \mid \mathbf{r})$ is as in \eqref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k(\mathbf{x}_{\mathbf{r},i},
\mathbf{x}_{\mathbf{u},j})$, where $\mathbf{x}_{\mathbf{r},i}$ and
$\mathbf{x}_{\mathbf{u},j}$ denote feature vectors for the $i$th state in
$\mathcal{S}$ and the $j$th state in $\mathbf{X_u}$, respectively
\cite{DBLP:conf/uai/JinDAS17}.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of parameters
$\bm\lambda$, inducing rewards $\mathbf{u}$, and the rewards for all relevant
states $\mathbf{r}$. While the previous paper to consider this IRL model
computed maximum likelihood (ML) estimates for $\bm\lambda$ and $\mathbf{u}$,
and made an assumption that $\mathbf{r}$ in \eqref{eq:r} has zero variance
\cite{DBLP:conf/nips/LevinePK11}, we aim to avoid this assumption and use
VI to approximate the full posterior distribution $\posterior$.
\emph{Variational inference} is an approximation technique for probability
densities \cite{blei2017variational}. Let $\approximation$ be our approximating
family of probability distributions for $\posterior$. Then the job of VI is to
optimise the parameters of the approximating distribution in order to minimise
the \emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation.  KL divergence (asymmetrically) measures how
different the two distributions are, and can be defined as
\cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant w.r.t. $\approximation$
\cite{blei2017variational}, so we can remove it from our optimisation objective.
The negation of what remains is often called the \emph{evidence lower bound}
(ELBO) and is defined as\footnote{Throughout the proposal, all integrals should
  be interpreted as definite integrals over the entire sample space.}
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iiint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}

By considering full probability distributions instead of point estimates---as
long as the approximations are able to capture important features of the
posterior---our predictions are likely to be more accurate and rely on fewer
assumptions. Moreover, we hope to make use of various recent advancements in VI
for both time complexity and approximation distribution fit, making the
resulting algorithm competitive both in terms of running time and model fit.

@@

\section{The Model} \label{sec:model}

>>>>

In this section, we introduce definitions and mathematical details relevant to
the problem. We begin by formally defining what the problem is.

@@

The optimal (deterministic) \emph{policy} $\pi : \mathcal{S} \to \mathcal{A}$
(i.e., a choice of actions for each state that maximises reward over time) is
usually constructed by defining a \emph{value (utility) function} $\V :
\mathcal{S} \to \mathbb{R}$ that measures how good a state is based on the
reward $\mathbf{r}$ as well as the structure of the MDP. One can then find $\V$
by applying the Bellman backup operator until convergence to every $s \in
\mathcal{S}$ (the technique is known as \emph{value iteration})
\cite{DBLP:books/daglib/0023820}:
\[
  V_{\mathbf{r}}(s) \coloneqq r(s) + \gamma \max_{a \in \mathcal{A}} \sum_{s' \in
    \mathcal{S}} \mathcal{T}(s, a, s')V_{\mathbf{r}}(s').
\]

However, we follow previous work on GP IRL
\cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}, and use a
\emph{linearly solvable} (or \emph{maximum causal entropy}) MDP with a
stochastic policy that defines probability distributions over actions (instead
of suggesting a single action for each state) \cite{ziebart2008maximum}. This
type of MDP can be solved by applying the `soft' version of the operator
\cite{DBLP:conf/nips/LevinePK11,supplementary_material}:
\begin{equation} \label{eq:update_rule}
  \V(s) \coloneqq \log \sum_{a \in \mathcal{A}} \exp\left( r(s) + \gamma\sum_{s'
      \in \mathcal{S}} \mathcal{T}(s, a, s')\V(s') \right).
\end{equation}
With this model, we can express the likelihood as
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}
\begin{align*}
  p(\mathcal{D} \mid \mathbf{r}) &= \prod_{i=1}^N \prod_{t=1}^T p(a_{i,t} \mid s_{i,t}) \\
                                 &= \exp\left( \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right),
\end{align*}
where
\[
  Q_{\mathbf{r}}(s, a) = r(s) + \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s, a, s')\V(s').
\]

% TODO: rewrite the 'we write' sentence and the sentence before that
As we want learned rewards to generalise to previously unseen states and
for states to have a notion of similarity, the IRL definition also includes $d$
features associated with each state. In this paper, we will focus on modelling
rewards as a function from feature space to $\mathbb{R}$ using a \emph{Gaussian
  process}. A GP is defined as a collection of random variables, any finite
number of which has a joint Gaussian distribution
\cite{DBLP:books/lib/RasmussenW06}. We write $r \sim \mathcal{GP}(0, k)$ to say
that $r$ is a GP with mean $0$ and covariance function $k$. \emph{Covariance
  functions} (also known as \emph{kernels}) take two state feature vectors as
input and quantify how similar the two states are, in a sense that we would
expect high covariance scores to be associated with similar rewards.

A common way to scale GPs to larger data sets is by selecting $m$ points in the
feature space--called \emph{inducing points}---and focus most of the training
effort on them \cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be
the $m \times d$ matrix of features at inducing points, and let $\mathbf{u}$ be
the rewards at those states. Then the full joint probability distribution can be
factorised as
\begin{equation} \label{eq:full}
  \pfull = p(\mathbf{u}) \times p(\mathbf{r} \mid \mathbf{u}) \times
  p(\mathcal{D} \mid \mathbf{r}),
\end{equation}
where
\begin{align*}
  p(\mathbf{u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \\
                &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \\
                &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right)
\end{align*}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}. The GP posterior is then a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11} defined as
\begin{equation} \label{eq:r}
  p(\mathbf{r} \mid \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru).
\end{equation}
The matrices such as
$\Kru$ are called \emph{covariance matrices} (also \emph{kernel matrices} and
\emph{Gram matrices}) and are defined as
$[\Kru]_{i,j} = k(\mathbf{x}_{\mathbf{r},i},
\mathbf{x}_{\mathbf{u},j})$, where $\mathbf{x}_{\mathbf{r},i}$ and
$\mathbf{x}_{\mathbf{u},j}$ denote feature vectors for the $i$th state in
$\mathcal{S}$ and the $j$th inducing point, respectively
\cite{DBLP:conf/uai/JinDAS17}.

Our goal is then to use VI in order to approximate $\posterior$.
Let $\approximation$ denote our approximation. VI aims to optimise this
approximation by minimising the \emph{Kullback-Leibler} (KL) divergence between
the original probability distribution and our approximation. KL divergence
(asymmetrically) measures the difference between two probability distributions
and can be defined as \cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}_{\approximation}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}_{\approximation}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}_{\approximation}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant with respect to
$\approximation$ \cite{blei2017variational}, so we can remove it from our
optimisation objective. The negation of what remains is known as the
\emph{evidence lower bound} (ELBO) and is defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iiint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}
Thus, instead of minimising KL divergence, we focus on maximising $\mathcal{L}$
by optimising the values of parameters to be defined in Section \ref{sec:model}.
Note that in this paper all integrals should be interpreted as definite
integrals over the entire sample space. Also, all expected values are with
respect to $\approximation$.

@@

\section{The Model} \label{sec:model}
% TODO: matrix notation should be at the start of The Problem
